A note on the instability and degeneracy of deep learning models

### 2016-11-29 writing/appendix/appendix-note.Rmd
edits from dr. nordman on note

{\bf Proof of Proposition \ref{prop:degen}.} Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min} \in \mathcal{X}^N$ as in the proof of Proposition \ref{prop:instab} where $|\mathcal{X}|<\infty$ holds in the FSFS model. We may suppose $|\mathcal{X}|>1$ (i.e., $\mathcal{X}^N$ has more than one outcome) because otherwise the model is trivially degenerate for all $N \geq 1$. Fix $0 < \epsilon < 1$ Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0$. If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition $P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
\begin{align*}
1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
& \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
\end{align*}
From the lower bound on $P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$ and the upper bound on $1-P_{\boldsymbol \theta_N}(\boldsymbol x_{max})$, it follows that
\begin{align*}
\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
\end{align*}
as $N \rightarrow \infty$ by the definition of an unstable FSFS model (c.f. Definition \ref{defn:instab}). Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.$\Box$

>>>>

{\bf Proof of Proposition \ref{prop:degen}.} Define $\boldsymbol x_{max}$ and $\boldsymbol x_{min} \in \mathcal{X}^N$ as in the proof of Proposition \ref{prop:instab} where $|\mathcal{X}|<\infty$ holds in the FSFS model. We may suppose $|\mathcal{X}|>1$ (i.e., $\mathcal{X}^N$ has more than one outcome) because otherwise the model is trivially degenerate for all $N \geq 1$. Fix $0 < \epsilon < 1$. Then, $\boldsymbol x_{max} \in M_{\epsilon, \boldsymbol \theta_N}$, so $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \ge P_{\boldsymbol \theta_N}(\boldsymbol x_{max}) > 0$. If $\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}$, then by definition $P_{\boldsymbol \theta_N}(\boldsymbol x) \le [P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}$ holds so that
\begin{align*}
1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})
& = \sum\limits_{\boldsymbol x \in \mathcal{X}^N \setminus M_{\epsilon, \boldsymbol \theta_N}}P_{\boldsymbol \theta_N}(\boldsymbol x) \\
& \le (|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}.
\end{align*}
From the lower bound on $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})$ and the upper bound on $1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})$, it follows that
\begin{align*}
\frac{1}{N}\log\left[\frac{P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}{1-P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N})}\right] & \ge \frac{1}{N} \log\left[\frac{P_{\boldsymbol\theta_N}(\boldsymbol x_{max})}{(|\mathcal{X}|^N)[P_{\boldsymbol \theta_N}(\boldsymbol x_{max})]^{1-\epsilon}[P_{\boldsymbol \theta_N}(\boldsymbol x_{min})]^{\epsilon}}\right] \\
&= \frac{\epsilon}{N} \log\left[\frac{P_{\boldsymbol \theta_N}(\boldsymbol x_{max})}{P_{\boldsymbol \theta_N}(\boldsymbol x_{min})}\right] - \log |\mathcal{X}| \rightarrow \infty
\end{align*}
as $N \rightarrow \infty$ by the definition of an unstable FSFS model (c.f. Definition \ref{defn:instab}). Consequently, $P_{\boldsymbol \theta_N}(M_{\epsilon, \boldsymbol \theta_N}) \rightarrow 1$ as $N \rightarrow \infty$ as claimed.$\Box$

### 2016-11-29 writing/note.Rmd
edits from dr. nordman on note

---
title: "A note on the instability and degeneracy of deep learning models"
author: | 
  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
  | Iowa State University
  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
date: "November, 2016"
linestretch: 2
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlated data structures found in several application areas, there is increasing interest in predicting/identifying instability. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes. (A sequence of probability models results indexed by $N$ accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$. In this case, a one component change in the data sequence can provably shift probability by orders of magnitude. Also, as a measure of instability becomes more extreme, the resulting probability models are shown  to tend to degeneracy, placing all their probability on arbitrarily small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
---

@@

We consider the behavior, and the potential impropriety, of probability models built to incorporate a sequence of discrete observations with length $N$. Let $(X_1, \dots, X_N)$ denote a set of sequence of discrete random variables with a finite sample space, $\mathcal{X}^N$. That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ takes values in the $N$-fold product space $\mathcal{X}^N$. For each $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$. We assume that the model support of $P_{\boldsymbol \theta_N}$ is the sample space $\mathcal{X}^N$. This framework produces a series $P_{\boldsymbol \theta_N}$ of probability models, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data of each length $N \geq 1$. (The size and structure of such parameters are without restriction, and natural cases include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.) We will call this model class *Finitely Supported Finite Sequence (FSFS) models*.

@@

For $\boldsymbol X = X_1, \dots, X_N$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function of the form
$$
p_{\boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
$$ 
for fixed positive integers $k$ and $L$ denoting the dimensions of the parameter and natural parameters spaces respectively, $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$, $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ a vector of sufficient statistics, 
$$
\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
$$
the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.

Defining $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models. Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables. For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).  This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}. @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\theta_N=\lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [cf. @smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$. For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ is a matrix with dimension $N_H \times N$, the RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
$$
p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
$$
where 
$$
\psi(\boldsymbol \theta) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta \in \Theta,
$$ 
is the normalizing function. Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM.

The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing this joint specification:
$$
p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} p_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
$$

Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accomodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hidens may also potentially increase with $N$. Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model. This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.

@@

**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and maintaining the restriction of only allowing conditional dependence between neighboring layers. The joint probability mass function for a DBM is
$$
p_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right], 
$$
for $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}$ where
$$
\psi(\boldsymbol \theta) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
$$ 
is the normalizing function and parameters in the model are $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$. Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.

The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification as
$$
p_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} p_{\boldsymbol \theta} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
$$

@@

Additionally, S-unstable FSFS model sequences are connected to degenerate models, where model *degeneracy* implies placing all probability on a small portion of the sample space. For perspective, note that differing bounds on $1/N\cdot\text{ELPR}(\boldsymbol \theta_N$ in (\ref{eqn:elpr}) provide a spectrum of levels of "stability" and Proposition \ref{prop:instab} indicates increasing and undesirable sensitivity of model probabilities (i.e., for one component changes in outcomes) as (\ref{eqn:elpr}) increases. Furthermore, as the instability measure (\ref{eqn:elpr}) grows, FSFS model sequences are guaranteed to slide into full degeneracy as Proposition \ref{prop:degen} will show. Define a $\epsilon$-modal set 
$$
M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) + \epsilon\min\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) \right\}
$$
of possible outcomes, for a given $0 < \epsilon < 1$.

>>>>

---
title: "A note on the instability and degeneracy of deep learning models"
author: | 
  | Andee Kaplan, Daniel Nordman, and Stephen Vardeman
  | Iowa State University
  | ajkaplan@iastate.edu, dnordman@iastate.edu, vardeman@iastate.edu
date: "November, 2016"
linestretch: 2
output: 
  pdf_document:
    keep_tex: true
    fig_caption: yes
    includes:
      in_header: latex/header.tex
    number_sections: yes
bibliography: references/refs.bib
abstract: |
  A probability model exhibits instability if small changes in a data outcome result in large, and often unanticipated, changes in probability. For correlated data structures found in several application areas, there is increasing interest in predicting/identifying instability. We consider the problem of quantifying instability for general probability models defined on sequences of observations, where each sequence of length $N$ has a finite number of possible outcomes. (A sequence of probability models results indexed by $N$ that accommodates data of expanding dimension.) Model instability is formally shown to occur when a certain log-probability ratio under such models grows faster than $N$. In this case, a one component change in the data sequence can provably shift probability by orders of magnitude. Also, as a measure of instability becomes more extreme, the resulting probability models are shown  to tend to degeneracy, placing all their probability on arbitrarily small portions of the sample space. These results on instability apply to large classes of models commonly used in random graphs, network analysis, and machine learning contexts.
---

@@

We consider the behavior, and the potential impropriety, of probability models built to incorporate a sequence of discrete observations with length $N$. Let $(X_1, \dots, X_N)$ denote a set of discrete random variables with a finite sample space, $\mathcal{X}^N$. That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ represents a finite set of potential outcomes for the variable $X_i$, and the data sequence $(X_1,\ldots,X_N)$ takes values in the $N$-fold product space $\mathcal{X}^N$. For each $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$. We assume that the model support of $P_{\boldsymbol \theta_N}$ is the sample space $\mathcal{X}^N$. This framework produces a series $P_{\boldsymbol \theta_N}$ of probability models, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data of each length $N \geq 1$. (The size and structure of such parameters are without restriction, and natural cases include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.) We will call this model class *Finitely Supported Finite Sequence (FSFS) models*.

@@

For $\boldsymbol X = (X_1, \dots, X_N)$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function of the form
$$
p_{N, \boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
$$ 
for fixed positive integers $k$ and $L$ denoting the dimensions of the parameter, $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$ and natural parameter $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ spaces, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ a vector of sufficient statistics, 
$$
\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
$$
the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.

Defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \lambda_N}(\boldsymbol x)$ with $\boldsmbol \theta}_N=\boldsymbol \lambda_N$ as $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models. Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables. (Note for random graphs or networks with, say, $m$ nodes, one may wish to consider $N={m \choose 2}$ edges as binary (presence/absence) variables $X_i$; in this case, the length $N$ of data sequence may naturally increase as a function of $m$.) For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).  This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}. @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\boldsymbol \theta_N=\boldsymbol \lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, binary variables being most common [cf. @smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$. For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ as a matrix with dimension $N_H \times N$, the RBM for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
$$
P_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
$$
where 
$$
\psi(\boldsymbol \theta) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta \in \Theta,
$$ 
is the normalizing function. Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM.

The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing this joint specification:
$$
P_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
$$

Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accomodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well; in particular, the number $N_H$ of hiddens may also potentially increase with $N$. Additionally, as $|\mathcal{X}| = 2$ and $p_{\boldsymbol \theta}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model. This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.

@@

**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and maintaining the restriction of only allowing conditional dependence between neighboring layers. The joint probability mass function for a DBM is
$$
P_{\boldsymbol \theta} (\tilde{\boldsymbol x}) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\boldsymbol \theta) \right], 
$$
for $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}$ where
$$
\psi(\boldsymbol \theta) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right], \qquad \boldsymbol \theta \in \Theta
$$ 
is the normalizing function and parameters in the model are $\boldsymbol \beta \in \mathbb{R}^V$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$ for $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension $N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$. Let $\boldsymbol \theta = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta \subset \mathbb{R}^{q(N)}$ denote the combined vector of parameters with total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$.

The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification as
$$
P_{\boldsymbol \theta} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} p_{\boldsymbol \theta} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
$$

@@

Additionally, S-unstable FSFS model sequences are connected to degenerate models, where model *degeneracy* implies placing all probability on a small portion of the sample space. For perspective, note that differing bounds on $1/N\cdot\text{ELPR}(\boldsymbol \theta_N)$ in (\ref{eqn:elpr}) provide a spectrum of levels of "stability" and Proposition \ref{prop:instab} indicates increasing and undesirable sensitivity of model probabilities (i.e., for one component changes in outcomes) as (\ref{eqn:elpr}) increases. Furthermore, as the instability measure (\ref{eqn:elpr}) grows, FSFS model sequences are guaranteed to slide into full degeneracy as Proposition \ref{prop:degen} will show. Define a $\epsilon$-modal set 
$$
M_{\epsilon, \boldsymbol \theta_N} \equiv \left\{\boldsymbol x \in \mathcal{X}^N: \log P_{\boldsymbol \theta_N}(\boldsymbol x) > (1-\epsilon)\max\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) + \epsilon\min\limits_{\boldsymbol x^* \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x^*) \right\}
$$
of possible outcomes, for a given $0 < \epsilon < 1$.
